---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pdfium

<!-- badges: start -->
[![R-CMD-check](https://github.com/humanpred/rpdfium/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/humanpred/rpdfium/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/humanpred/rpdfium/branch/main/graph/badge.svg)](https://app.codecov.io/gh/humanpred/rpdfium)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/pdfium)](https://CRAN.R-project.org/package=pdfium)
[![Codecov test coverage](https://codecov.io/gh/humanpred/rpdfium/graph/badge.svg)](https://app.codecov.io/gh/humanpred/rpdfium)
<!-- badges: end -->

`pdfium` provides idiomatic R bindings to
[Google's PDFium engine](https://pdfium.googlesource.com/pdfium/) — the
same library that powers Chrome's PDF viewer. It has two halves:

* a **read surface** that exposes vector-path geometry —
  stroke / fill / Bezier control points / transformation matrices —
  alongside text, fonts, images, annotations, form fields,
  attachments, signatures, structure tree, and rendering. The path
  geometry, in particular, no other CRAN package surfaces today.
* a **mutation surface** (opt-in via `readwrite = TRUE`) that lets
  you rotate / reorder / merge pages, draw fresh page objects,
  create and edit annotations, fill form fields, and add file
  attachments — then save the result.

## What it is for

* **Auditing** PDF figures (which lines, which colors, which fonts).
* **Extracting** curves from regulatory filings and scientific
  publications.
* **Building** PDF normalization pipelines that need geometry, not
  just text.
* **Filling** AcroForm fields programmatically and flattening the
  result for downstream tooling.
* **Authoring** programmatic PDFs from vector graphics, JPEG
  images, text in the 14 standard fonts or any TrueType / Type1
  typeface, and annotations (think: figure callouts, table
  reports, annotated source documents). `/Info`-dict writes and
  on-save encryption are the remaining v0.1.0 gaps — both need
  upstream PDFium changes that we've proposed but Google hasn't
  shipped yet.
* Anything you'd otherwise drop into Python with `pypdfium2`.

See [`vignette("mutating-pdfs")`](https://humanpred.github.io/rpdfium/articles/mutating-pdfs.html)
for a walkthrough of the writer surface, and
[`vignette("comparison")`](https://humanpred.github.io/rpdfium/articles/comparison.html)
for how `pdfium` lines up against `pdftools`, `qpdf`, `magick`,
`tabulizer`, and `staplr`.

## Status

First CRAN release (`0.1.0`). The public API is documented on the
[pkgdown site](https://humanpred.github.io/rpdfium/) and exercised at
100% R coverage; architectural decisions for the release are recorded
under `dev/decisions/`.

## Installation

`pdfium` downloads its `libpdfium` binary from
[bblanchon/pdfium-binaries](https://github.com/bblanchon/pdfium-binaries)
at install time. The pinned version lives in
`tools/pdfium-version.txt`. If your install runs without internet
access, set `PDFIUM_OFFLINE=1` and place the matching tarball under
`inst/pdfium-binaries/` before installing.

```r
# Release version (once on CRAN):
install.packages("pdfium")

# Development version:
remotes::install_github("humanpred/rpdfium")
```

## Example

```{r example, eval = FALSE}
library(pdfium)

doc <- pdf_doc_open(system.file("extdata", "fixtures", "minimal.pdf",
  package = "pdfium"
))
pdf_page_count(doc)
pdf_doc_close(doc)
```

More examples ship in the vignettes (`vignette("getting-started",
package = "pdfium")`, etc.) and on the
[pkgdown site](https://humanpred.github.io/rpdfium/).

## License

`pdfium` is MIT-licensed. The bundled `libpdfium` binary is BSD-3-Clause
and is *not* distributed in the source tarball — see
[`LICENSE.md`](LICENSE.md) and
[`dev/decisions/ADR-003-binary-distribution.md`](dev/decisions/ADR-003-binary-distribution.md).
