---
title: "How pdfium is built"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How pdfium is built}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette explains the four-layer model `pdfium` uses, the memory
contract you rely on as a user, and the most common pitfalls. For
contributor-facing detail (CI topology, fixture rebuild pipeline,
PDFium bump procedure), see `docs/architecture.md` in the source tree.

## Four-layer model

```
R user
  │
  ▼
R API           pdf_open(), pdf_close(), pdf_page_count(), ...
  │             S3 classes: pdfium_doc, pdfium_page, pdfium_obj
  ▼
Rcpp glue       internal cpp_* helpers; you should never see them
  │
  ▼
PDFium C ABI    Google's PDFium engine (BSD-3-Clause)
  │
  ▼
libpdfium       shared library downloaded at install time
```

The R API takes user inputs, validates them, calls the Rcpp glue, and
shapes the result back into idiomatic R (tibbles for tabular outputs,
S3 objects for handles). You only ever interact with the R API.

## Memory: what happens when you forget to close

```{r, eval = FALSE}
inspect <- function(path) {
  doc <- pdf_open(path)
  pdf_page_count(doc)
  # `doc` goes out of scope here.
  # R's GC will eventually finalize it and call FPDF_CloseDocument.
}
inspect("report.pdf")
```

This is safe. `pdfium` registers a C finalizer on every PDF handle.
When the R object becomes unreachable, R's garbage collector reclaims
it; the finalizer then calls PDFium's `FPDF_CloseDocument` to release
the underlying memory.

The caveat is that GC is **eventual**, not deterministic. If you open
many large documents in a tight loop, you may exhaust process memory
before the GC catches up. In that case, close explicitly:

```{r, eval = FALSE}
inspect <- function(path) {
  doc <- pdf_open(path)
  on.exit(pdf_close(doc), add = TRUE)
  pdf_page_count(doc)
}
```

`pdf_close()` is **idempotent**: calling it twice is a no-op. The
finalizer notices the handle has already been closed and skips its
own close call. You can safely combine explicit close with the
automatic fallback.

## Children outlive their parent

```{r, eval = FALSE}
load_page <- function(path) {
  doc <- pdf_open(path)
  page <- pdf_load_page(doc, 1)   # available in Phase 1+
  pdf_close(doc)                  # this is fine
  page                            # still usable here
}
```

When you call `pdf_load_page(doc, ...)`, the returned `pdfium_page`
holds an internal reference to its parent `pdfium_doc`. Even if you
drop your reference to `doc` (or explicitly close it), the page stays
valid until the page object itself is collected. The underlying PDFium
document is kept alive in the background until the last page (or
object) that depends on it goes away.

The order is: **child references the parent**, never the other way
around.

## How the binary gets loaded

When you `library(pdfium)`, this happens:

1. R loads the package's compiled `pdfium.so` (or `pdfium.dll` on
   Windows).
2. The dynamic linker follows the RPATH baked in at install time and
   loads `libpdfium.{so|dylib|dll}` from the package's `inst/lib/`.
3. `.onLoad` calls `FPDF_InitLibraryWithConfig()` exactly once.
4. When you call `library.unload("pdfium")` or quit R, `.onUnload`
   runs `FPDF_DestroyLibrary()`.

The `libpdfium` binary was downloaded the first time you installed
`pdfium`. The pinned release tag lives in `tools/pdfium-version.txt`.
If your machine has no internet access at install time, set
`PDFIUM_OFFLINE=1` and place the matching tarball under
`inst/pdfium-binaries/` before installing.

## Common gotchas

- **Holding many documents at once.** GC is non-deterministic; close
  explicitly with `pdf_close()`.
- **Deleting an open PDF on Windows.** Windows blocks deletion of
  files held by open handles. Close the document first, then delete.
- **Using a closed handle.** Functions that take a `pdfium_doc` raise
  an error if the handle has already been closed. Re-open the file
  if you need it again.
- **Calling `FPDF_*` directly.** Don't. The R API exists for a reason
  — bypassing it bypasses the lifetime tracking and you'll crash R.

## Further reading

- The decision records under `docs/decisions/` capture every
  architectural choice and the alternatives that were considered.
- `docs/architecture.md` covers contributor-facing topics: CI
  topology, the PDFium bump procedure, and the fixture-rebuild
  pipeline.
- `vignette("getting-started")` walks through a complete inspection
  workflow.
