---
title: "Extracting path geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting path geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Vector path geometry is the gap `pdfium` fills relative to other R PDF
libraries. `pdftools` and friends give you the rendered raster and the
text content but not the underlying path commands; this vignette walks
through the path API end-to-end.

```{r}
library(pdfium)
fixture <- system.file("extdata", "fixtures", "shapes.pdf",
  package = "pdfium"
)
```

## Per-page-object access

The lowest-level entry point is `pdf_path_segments(obj)` on a single
`"path"`-typed `pdfium_obj`. Each row is one path segment:

```{r}
doc <- pdf_doc_open(fixture)
page <- pdf_page_load(doc, 1L)
paths <- Filter(function(o) o$type == "path", pdf_page_objects(page))
length(paths)

pdf_path_segments(paths[[1L]])
```

Columns:

* `segment_index` — 1-based segment index within this path.
* `segment_type` — one of `"moveto"`, `"lineto"`, `"bezierto"`, or
  `"unknown"`.
* `x`, `y` — segment coordinates in PDF user space (points, origin
  bottom-left).
* `close_figure` — `TRUE` on the final segment of a closed sub-path.

A note on Bezier curves: PDFium stores a cubic curve as three consecutive
`"bezierto"` rows — the two control points followed by the endpoint. The
v0.1.0 readout returns each of those three as separate rows. A
companion accessor `pdf_path_bezier_controls()` that returns the pair
of control points alongside the endpoint is gated on an upstream PDFium
patch ([CL 147810](https://pdfium-review.googlesource.com/c/pdfium/+/147810));
see [ADR-009](https://github.com/humanpred/rpdfium/blob/main/dev/decisions/ADR-009-defer-bezier-controls.md)
for the full rationale and status.

## Path style

Stroke and fill colour come back as 4-element numeric vectors `(red,
green, blue, alpha)` with values in 0..255. NA's indicate "no stroke" or
"no fill" — paths can have either, both, or neither:

```{r}
pdf_path_stroke(paths[[1L]])
pdf_path_fill(paths[[1L]])
```

Dash pattern is a tibble with `pattern` (the on-off lengths) and
`phase` (the offset into the pattern):

```{r}
pdf_path_dash(paths[[1L]])
```

The path's transformation matrix (the CTM applied to its local
coordinates) comes from `pdf_obj_matrix()`:

```{r}
pdf_obj_matrix(paths[[1L]])
```

## One-call extraction

For consumers like `kmextract` that want one row per segment across a
page with style folded in, `pdf_extract_paths()` is the batched API:

```{r}
all_paths <- pdf_extract_paths(fixture)
all_paths
```

The result is a tibble. Schema:

* `path_index` — 1-based path-object index within the page (paths
  only; non-path objects don't appear).
* `segment_index`, `segment_type`, `x`, `y`, `close_figure` —
  same content as `pdf_path_segments()`.
* `stroke_red/green/blue/alpha`, `fill_red/green/blue/alpha` —
  stroke and fill colours, 0..255. `NA` for paths with no stroke /
  no fill.
* `stroke_width` — stroke width in user space; `NA` for unstroked
  paths.
* `bounds_left/bottom/right/top` — path's bounding box in PDF user
  space.

Three attributes carry per-page context:

```{r}
attr(all_paths, "page_size")
attr(all_paths, "page_rotation")
nrow(attr(all_paths, "text_runs"))
```

* `page_size` — named numeric vector of `width` and `height` in
  PDF points.
* `page_rotation` — integer 0/90/180/270 (the page's own
  `/Rotate` attribute).
* `text_runs` — a tibble of every text run on the page, shaped like
  the output of `pdf_text_runs()`. Convenient when downstream consumers
  want to position labels relative to a curve.

## Filtering and inspecting paths in practice

A common workflow: select paths by stroke colour and inspect the segment
endpoints.

```{r}
red_paths <- all_paths[
  !is.na(all_paths$stroke_red) &
    all_paths$stroke_red > 200 &
    all_paths$stroke_green < 50,
]
nrow(red_paths)
range(red_paths$x)
range(red_paths$y)
```

## Cleanup

```{r}
pdf_page_close(page)
pdf_doc_close(doc)
```
