---
title: "Getting started with pdfium"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pdfium}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pdfium` exposes Google's [PDFium](https://pdfium.googlesource.com/pdfium/)
engine to R. Where `pdftools` (Poppler-based) gives you the text content
and a rasterised page, `pdfium` adds vector-level structure: every path
object's segments and style, every text run's font and bounding box, every
embedded image's source bytes and metadata, every form XObject's
nested children, and every clip region's geometry. It complements rather
than replaces `pdftools` — the two cover different parts of the PDF inspection
surface.

This vignette walks through opening a document, inspecting its structure,
and pulling out a few common things. Three later vignettes go deeper into
[paths](extracting-paths.html), [text](text-extraction.html), and
[rendering](rendering.html).

```{r}
library(pdfium)
```

## Opening and closing a document

```{r}
fixture <- system.file("extdata", "fixtures", "shapes.pdf",
  package = "pdfium"
)
doc <- pdf_doc_open(fixture)
doc
```

The returned object is an S3 handle wrapping a PDFium `FPDF_DOCUMENT`. R's
garbage collector calls `FPDF_CloseDocument` when the handle becomes
unreachable, so you don't *have* to close it explicitly — but for large
documents or when you need to delete the source file afterward (Windows
blocks deletion of open files), close it yourself with `pdf_doc_close()`:

```{r}
pdf_count <- pdf_page_count(doc)
pdf_count
```

Both `pdf_page_count()` and `pdf_doc_info()` also accept a path directly
when you want a one-shot inspection. The package opens and closes the
document for you in that case:

```{r}
pdf_page_count(fixture)
```

## Document metadata

`pdf_doc_info()` returns the page count, file version, every standard
Info-dictionary entry, and POSIXct parses of the two date fields. The
shape mirrors `pdftools::pdf_info()` for easy porting:

```{r}
info <- pdf_doc_info(doc)
info$page_count
info$file_version # PDFium reports 10 * major + minor (17 = PDF 1.7)
info$producer
info$creation_date_parsed
```

For a single tag use `pdf_doc_meta(doc, "Producer")`. The standard tags
PDFium recognises are `"Title"`, `"Author"`, `"Subject"`, `"Keywords"`,
`"Creator"`, `"Producer"`, `"CreationDate"`, `"ModDate"`, and
`"Trapped"`. Missing tags return the empty string.

If you have a date string in PDF format (`"D:YYYYMMDDHHmmSS+HH'mm'"`),
`pdf_parse_date()` parses it into UTC POSIXct. It's vectorised and
accepts truncated forms:

```{r}
pdf_parse_date(c("D:20240115123045Z", "D:2024"))
```

## Pages and page objects

A page is loaded with `pdf_page_load(doc, page_num)`. The returned handle
carries a reference to its parent doc so GC ordering is safe:

```{r}
page <- pdf_page_load(doc, 1L)
pdf_page_size(doc, 1L) # width and height in PDF points
pdf_page_rotation(doc, 1L) # 0, 90, 180, or 270 degrees
```

Every drawable element on a page is a *page object*. `pdf_page_objects()`
returns them as a list of typed `pdfium_obj` handles:

```{r}
objs <- pdf_page_objects(page)
length(objs)
vapply(objs, function(o) o$type, character(1))
```

Each object's type is one of `"path"` (vector path with line/curve
segments), `"text"` (a text run), `"image"` (embedded raster), `"form"`
(a Form XObject - a reusable sub-page; see the section below), or
`"shading"` (a smooth-shading object).

You can read each object's bounding box on the page in PDF user space
(points, origin at the page's bottom-left):

```{r}
pdf_obj_bounds(objs[[1L]])
```

And its transformation matrix (the CTM applied when the object is drawn):

```{r}
pdf_obj_matrix(objs[[1L]])
```

The accessors that work on every page object are `pdf_obj_type()`,
`pdf_obj_bounds()`, `pdf_obj_matrix()`. Type-specific accessors
follow naming conventions: `pdf_path_*`, `pdf_text_*`, `pdf_image_*`.

## Path geometry

For path objects, `pdf_path_segments()` returns a tibble of segment
coordinates plus a `close_figure` flag on each segment:

```{r}
paths <- Filter(function(o) o$type == "path", objs)
pdf_path_segments(paths[[1L]])
```

Stroke and fill colour, dash pattern, and miscellaneous style come back as
separate accessors:

```{r}
pdf_path_stroke(paths[[1L]])
pdf_path_fill(paths[[1L]])
pdf_path_dash(paths[[1L]])
```

For a one-call extraction that returns one row per segment across every
path on every page — the shape `kmextract` and similar consumers expect — use
`pdf_extract_paths(path_or_doc)`:

```{r}
extracted <- pdf_extract_paths(fixture)
str(extracted, max.level = 1L)
nrow(extracted)
head(extracted, 3L)
```

The result is a tibble with one row per path segment, plus per-path
attributes (stroke/fill color, page size, etc.) and a `text_runs`
attribute carrying the page's text in tabular form. See the
[extracting-paths](extracting-paths.html) vignette for the full schema.

## Text content

`pdf_text_content(text_obj)` returns the Unicode string for one text
object, and `pdf_text_runs(page)` enumerates every text run on a page
with positions, font sizes, and font metadata:

```{r}
runs <- pdf_text_runs(page)
runs
```

For one-text-object access (font name, weight, embedded flag) use
`pdf_text_font(text_obj)`. See the
[text-extraction](text-extraction.html) vignette for the full set.

## Rendering

To rasterise a page, `pdf_render_page(page_or_doc, dpi)` returns a
`pdfium_bitmap` object that inherits from base R's `nativeRaster`:

```{r}
bmp <- pdf_render_page(doc, dpi = 96)
bmp
dim(bmp) # height, width
```

`pdfium_bitmap` is directly usable by `graphics::plot()`,
`graphics::rasterImage()`, and `grid::rasterGrob()`. Three converters
produce other common shapes:

```{r}
arr <- as.array(bmp) # 3D [H, W, 4] doubles in 0..1, like png::writePNG
ras <- as.raster(bmp) # base R "raster" class, hex-color matrix
mat <- as.matrix(bmp) # plain character matrix
```

The save helper `pdf_render_to_png(file, ...)` writes the bitmap straight
to a PNG file via the `png` package (a Suggests dependency).

## Embedded images and form XObjects

For object-type-specific extraction, see:

- [`pdf_image_info()`](../reference/pdf_image_info.html) /
  [`pdf_image_bitmap()`](../reference/pdf_image_bitmap.html) /
  [`pdf_image_data()`](../reference/pdf_image_data.html) for
  embedded raster images. The raw-stream-bytes path
  (`pdf_image_data(decoded = FALSE)`) lets you save the original
  JPEG/JBIG2/JPEG2000 without re-encoding.
- [`pdf_form_objects()`](../reference/pdf_form_objects.html) to walk
  the page objects nested inside a Form XObject (a reusable sub-page
  used for tiling patterns, glyph procedures, annotation appearance
  streams, etc.).
- [`pdf_obj_clip_path()`](../reference/pdf_obj_clip_path.html) /
  [`pdf_clip_path_segments()`](../reference/pdf_clip_path_segments.html)
  to read the clip geometry attached to a page object.

## Mutating a PDF

Everything above is read-only. To change a document — rotate pages,
restyle objects, fill form fields, add annotations, attach files —
open with `readwrite = TRUE` and save with `pdf_save()`:

```{r, eval = FALSE}
src <- system.file("extdata", "fixtures", "annotated.pdf",
                   package = "pdfium")
doc <- pdf_doc_open(src, readwrite = TRUE)

# Fill the textfield "name" with "Ada".
fields <- pdf_form_fields(doc)
name <- fields[vapply(fields, pdf_form_field_name, "") == "name"][[1]]
pdf_form_field_set_value(name, "Ada")

# Flatten the page so the value is baked into the content stream.
page <- pdf_page_load(doc, 1)
pdf_page_flatten(page)

# Persist.
pdf_save(doc, "filled.pdf")
pdf_doc_close(doc)
```

The full mutation surface is documented in the pkgdown reference
under "Structural mutation", "Page-object styling", "Path geometry",
"Annotation authoring", "Form filling", and "Attachment authoring".
`pdf_doc_new()` builds writable documents from scratch.

## Cleanup

```{r}
pdf_page_close(page)
pdf_doc_close(doc)
```

`pdf_doc_close()` and `pdf_page_close()` are idempotent — calling them more
than once is a no-op. R's GC also runs the finalisers automatically when
handles become unreachable. The [architecture](architecture.html) vignette
covers the memory model in depth.
