---
title: "Mutating PDFs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mutating PDFs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through the v0.1.0 mutation surface. Everything
here is opt-in: you have to either open a document with
`readwrite = TRUE`, or build one with `pdf_doc_new()`. Setters check
the readwrite flag and raise a clean error otherwise, so accidental
mutations on an inspection-only doc fail loudly.

```{r}
library(pdfium)
```

## The common pattern

```{r, eval = FALSE}
doc <- pdf_doc_open("report.pdf", readwrite = TRUE)
on.exit(pdf_doc_close(doc), add = TRUE)

# ... mutate ...

pdf_save(doc, "report-modified.pdf")
```

`pdf_save()` triggers `FPDFPage_GenerateContent()` on every page
that's been touched by a setter, so the content stream you serialise
reflects the changes. `pdf_save_to_raw(doc)` writes to a `raw`
vector instead of a file.

## Structural mutation

Move, rotate, resize, add, remove, and merge pages.

```{r, eval = FALSE}
# Rotate the first page 90 degrees clockwise.
pdf_page_set_rotation(doc, page_num = 1, rotation = 90)

# Set the crop box on page 2.
pdf_page_set_box(doc, page_num = 2,
                 box = "crop", bounds = c(36, 36, 576, 756))

# Reorder pages: move page 3 to position 1.
pdf_pages_reorder(doc, new_order = c(3, 1, 2, 4:pdf_page_count(doc)))

# Append a blank letter-size page.
pdf_page_new(doc, width = 612, height = 792)

# Merge another document's pages 1-5 onto the end.
other <- pdf_doc_open("appendix.pdf")
pdf_docs_merge(doc, other, source_pages = 1:5)
pdf_doc_close(other)

# 2-up imposition: print 4 logical pages onto each physical sheet.
pdf_n_up(doc, n_up = c(2, 2), output_size = c(612, 792))

# Tag the document's language for accessibility tooling.
pdf_doc_set_language(doc, "en-US")
```

## Building a page from scratch

`pdf_doc_new()` returns an empty writable document. Combine it with
`pdf_page_new()` and the page-object creators to draw programmatic
PDFs.

```{r, eval = FALSE}
doc <- pdf_doc_new()
page <- pdf_page_new(doc, width = 612, height = 792)

# A red filled rectangle (x, y, width, height) in PDF points.
rect <- pdf_rect_new(page, x = 50, y = 600, width = 200, height = 100)
pdf_path_set_fill(rect, color = c(220, 50, 47))
pdf_path_set_draw_mode(rect, fill_mode = "winding", stroke = FALSE)

# A text object on top.
pdf_text_new(page, "Hello pdfium!",
             font = "Helvetica-Bold", font_size = 18,
             x = 80, y = 640)

pdf_save(doc, "hello.pdf")
```

`pdf_path_new()` makes an empty path you grow with
`pdf_path_move_to()` / `pdf_path_line_to()` / `pdf_path_bezier_to()` /
`pdf_path_close()`. `pdf_path_append()` lets you pour a whole
data-frame of segments (matching the schema of `pdf_path_segments()`)
into a path in one call — useful for echoing geometry from one
document into another.

### Embedding a JPEG image

`pdf_image_new()` wraps `FPDFImageObj_LoadJpegFileInline` — the JPEG
bytes are copied into the PDF up front, so the source raw vector or
file is free immediately after the call. Pass `bounds = c(left,
bottom, right, top)` to place the image at a specific rectangle in
PDF user-space points; without `bounds`, the image sits at the
origin at its natural pixel size (rarely what you want).

```{r, eval = FALSE}
jpeg_path <- system.file("img", "Rlogo.jpg", package = "jpeg")
pdf_image_new(page, jpeg_path,
              bounds = c(72, 400, 272, 500))
```

PNG / TIFF / raw-bitmap embedding is deferred to a later release —
convert to JPEG with `magick::image_write(..., format = "jpeg")` if
needed.

### Custom fonts

The 14 PDF standard fonts work without embedding — pass the name
straight through (`pdf_text_new(page, "...", font = "Times-Italic",
...)`). For arbitrary TrueType or Type1 typefaces, load the font
first via `pdf_font_load()` (or `pdf_font_load_standard()` if you
just want a handle for a named standard font) and pass the
resulting `pdfium_font` handle into `pdf_text_new()`:

```{r, eval = FALSE}
font <- pdf_font_load(doc, "/path/to/MyTypeface-Regular.ttf")
pdf_text_new(page, "ünïcödé ✓", font = font, font_size = 24,
             x = 72, y = 300)
```

The font handle carries an `FPDFFont_Close` finalizer; the doc is
pinned in the externalptr's `prot` slot. `pdf_font_close(font)` is
idempotent and useful when you want deterministic release.

## Styling an existing object

The setters mirror the readers one-for-one:

```{r, eval = FALSE}
page <- pdf_page_load(doc, 1)
objs <- pdf_page_objects(page)

# Find the first path object.
i <- match(TRUE, vapply(objs, pdf_obj_type, "") == "path")
path <- objs[[i]]

# Make it dashed and turquoise.
pdf_path_set_dash(path, dash_array = c(4, 2), dash_phase = 0)
pdf_path_set_stroke(path, color = c(0, 200, 200))
```

Most styling setters are **composite**: they accept either a full
replacement (`color = c(r, g, b)` or `c(r, g, b, a)`) or a partial
overlay (individual `red` / `green` / `blue` / `alpha` arguments).
Integer 0-255 and double 0-1 colors are auto-detected. See
`?pdf_obj_set_blend_mode` for the full list of styling setters.

## Annotation authoring

```{r, eval = FALSE}
# Add a yellow highlight to page 1 covering a quad.
hl <- pdf_annot_new(page, subtype = "highlight",
                    bounds = c(100, 700, 400, 720))
pdf_annot_set_color(hl, color = c(255, 240, 0))
pdf_annot_set_contents(hl, "Important")
pdf_annot_append_quad(hl, quad = c(100, 700, 400, 700,
                                    100, 720, 400, 720))
```

The supported subtypes are listed in `?pdf_annot_new`. Delete an
annotation with `pdf_annot_delete(annot)`; the handle is invalidated
on the spot, so subsequent reads through it raise a clean error.

## Form filling

`pdf_form_field_set_value()` is polymorphic — it dispatches by the
field's type.

```{r, eval = FALSE}
fields <- pdf_form_fields(doc)
by_name <- setNames(fields,
                    vapply(fields, pdf_form_field_name, ""))

# Text field: character scalar.
pdf_form_field_set_value(by_name[["full_name"]], "Ada Lovelace")

# Checkbox: logical, or the literal export-value string.
pdf_form_field_set_value(by_name[["subscribe"]], TRUE)
pdf_form_field_set_value(by_name[["over_18"]], "Yes")

# Combobox / listbox: must match one of the field's options.
pdf_form_field_set_value(by_name[["country"]], "United Kingdom")
```

To wipe a single field back to its default value, use
`pdf_form_field_clear()`. To reset every field in the document,
`pdf_form_reset(doc)`.

When you're ready to ship a non-editable PDF, `pdf_page_flatten()`
bakes the widget appearances and any other annotations into the
page's content stream. **It's irreversible** — there is no
`pdf_page_unflatten()`. Run it as the last step before saving the
final-state PDF.

## Embedded file attachments

```{r, eval = FALSE}
# Add a CSV attachment to the document.
att <- pdf_attachment_new(doc, name = "data.csv")
csv_bytes <- charToRaw("a,b,c\n1,2,3\n")
pdf_attachment_set_data(att, csv_bytes)
pdf_attachment_set_dict_value(att, "Desc",
                              "Source data for the figures above")
```

Note the ordering: `pdf_attachment_set_data()` first (it creates
`/Params` on the embedded file stream as a side effect), then any
`pdf_attachment_set_dict_value()` calls.

## What's *not* in v0.1.0

A few writable surfaces are documented as upstream gaps awaiting a
PDFium patch:

- **Choice-field `/Opt` arrays.** `pdf_form_field_set_value()`
  validates against the existing options, but PDFium has no public
  API to grow the option list. See
  `dev/upstream-patches/pdfium-FPDFAnnot_AppendOption.patch`.
- **The file-stream `/Subtype` on attachments.** The MIME type
  surfaced by `pdf_attachment_mime_type()` lives on the embedded
  file stream itself, not on `/Params`; PDFium has no setter for it.
- **Document-info dict (`/Info`) writes** — title, author, subject.
  The reader (`pdf_doc_info()`) exists but the writer requires an
  upstream `FPDF_SetMetaText` that doesn't ship yet.

See `dev/upstream-patches/` in the source tree for the full set of
upstream patches we maintain.
