---
title: "Rendering pages to bitmaps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering pages to bitmaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pdf_render_page()` rasterises a PDF page through PDFium's renderer.
The result is a `pdfium_bitmap` S3 object that inherits from base R's
`nativeRaster`, so it plugs into `graphics::plot()`,
`graphics::rasterImage()`, and `grid::rasterGrob()` with no conversion.
Three converters cover the other common shapes downstream packages
expect.

```{r}
library(pdfium)
fixture <- system.file("extdata", "fixtures", "shapes.pdf",
  package = "pdfium"
)
```

## A first render

```{r, fig.width = 5, fig.height = 4}
doc <- pdf_open(fixture)
bmp <- pdf_render_page(doc, dpi = 96)
bmp # a one-line summary
dim(bmp) # height, width in pixels

plot(bmp) # uses plot.pdfium_bitmap()
```

The `plot()` call dispatches to `plot.pdfium_bitmap()`, which opens
a fresh plot window with `asp = 1` and zero margins and draws the
bitmap with [graphics::rasterImage()]. Pass `interpolate = FALSE`
when you want pixel-exact (nearest-neighbour) display of a small
bitmap — useful for embedded raster fixtures.

The bitmap's dimensions scale linearly with `dpi`. At the default
`dpi = 72`, one pixel per PDF point: a 4 × 3 inch page becomes 288 × 216
pixels. At `dpi = 144`, the same page is 576 × 432.

```{r}
dim(pdf_render_page(doc, dpi = 72))
dim(pdf_render_page(doc, dpi = 144))
```

## Background and transparency

The default background is white. Pass any colour string
`grDevices::col2rgb()` understands, or `NA` for transparent:

```{r}
bmp_red <- pdf_render_page(doc, dpi = 72, background = "red")
bmp_trans <- pdf_render_page(doc, dpi = 72, background = NA)

bmp_red[1L, 1L] # top-left pixel
bmp_trans[1L, 1L] # depends on whether page content covers
```

## Rotation

Rotation in degrees, applied on top of the page's own `/Rotate`
attribute. Rotating 90 or 270 swaps the bitmap's width and height:

```{r}
dim(pdf_render_page(doc, dpi = 72, rotation = 0))
dim(pdf_render_page(doc, dpi = 72, rotation = 90))
dim(pdf_render_page(doc, dpi = 72, rotation = 180))
dim(pdf_render_page(doc, dpi = 72, rotation = 270))
```

## Converting to other shapes

```{r}
# A 3D numeric array [H, W, 4] with values in 0..1 - matches the
# format png::writePNG() and pdftools::pdf_render_page() both produce.
arr <- as.array(bmp)
dim(arr)
range(arr)

# Base R "raster" object - character matrix of "#RRGGBBAA" hex colors.
ras <- as.raster(bmp)
ras[1L, 1L]

# A plain character matrix (drops the "raster" class).
mat <- as.matrix(bmp)
class(mat)
```

## Saving to PNG

For a one-call save, `pdf_render_to_png()` writes the rendered bitmap
to a PNG file. It needs the `png` package (a Suggests dependency):

```{r, eval = requireNamespace("png", quietly = TRUE)}
out <- tempfile(fileext = ".png")
pdf_render_to_png(doc, file = out, dpi = 96)
file.exists(out)
file.size(out)
```

## Annotations

`pdf_render_page(annotations = TRUE)` paints annotation appearance
streams on top of the page. shapes.pdf has no annotations so the flag
is a no-op here; on annotated PDFs the rendered bitmap visibly changes.

## Working with embedded images

For each `"image"`-typed page object, three accessors return rendered or
raw image data with the same `pdfium_bitmap` shape:

```{r}
img_fixture <- system.file("extdata", "fixtures", "image.pdf",
  package = "pdfium"
)
img_doc <- pdf_open(img_fixture)
img_page <- pdf_load_page(img_doc, 1L)
imgs <- Filter(function(o) o$type == "image", pdf_page_objects(img_page))

# Decoded source-pixel bitmap (no CTM applied).
src <- pdf_image_bitmap(imgs[[1L]])
dim(src)

# CTM-applied rendering (what a viewer would actually draw).
viewer <- pdf_image_rendered(imgs[[1L]])
dim(viewer)
```

For saving the original embedded asset without re-encoding, the raw-bytes
path is useful:

```{r}
filters <- pdf_image_filters(imgs[[1L]])
filters
raw_bytes <- pdf_image_data(imgs[[1L]], decoded = FALSE)
length(raw_bytes)
```

When `filters` contains `"DCTDecode"`, the raw bytes are the original
JPEG; when it contains `"JPXDecode"`, they're JPEG 2000; for
`"FlateDecode"` they're Deflate-compressed pixels. See
[`pdf_image_filters()`](../reference/pdf_image_filters.html) for the
full enumeration of common decoders.

## Cleanup

```{r}
pdf_close_page(img_page)
pdf_close(img_doc)
pdf_close(doc)
```
