---
title: "Extracting text and fonts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting text and fonts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pdfium` exposes text at the level of *text runs* — the contiguous runs
of glyphs that share a font, size, and rendering state. Each run carries
its bounding box, the Unicode content, and font metadata. Where
`pdftools::pdf_doc_text()` returns plain strings, `pdfium` lets you connect
each character to where it sits on the page and what font drew it.

```{r}
library(pdfium)
fixture <- system.file("extdata", "fixtures", "unicode.pdf",
  package = "pdfium"
)
doc <- pdf_doc_open(fixture)
page <- pdf_page_load(doc, 1L)
```

## Whole-page extraction

`pdf_text_runs(page)` returns one row per text-typed page object on the
page, with content + position + font in a single tibble:

```{r}
runs <- pdf_text_runs(page)
runs
```

Columns (use `names(runs)` for the canonical list — the names below
match the schema PDFium reports today):

* `object_index` — 1-based page-object index. Matches the index in
  `pdf_page_objects(page)`.
* `bbox_left`, `bbox_bottom`, `bbox_right`, `bbox_top` — bounding
  box in PDF user space (points, origin bottom-left).
* `font_size_pt` — the font size as PDFium reports it (in user-space
  points, before the text matrix is applied).
* `text` — Unicode content of the run, UTF-8 encoded.
* `font_base_name` — `BaseFont` from the font dictionary.
* `font_family` — `FontFamily` if present.
* `font_weight` — integer 100..900 (CSS-style; 400 ≈ regular,
  700 ≈ bold).
* `font_italic_angle` — italic angle in degrees, 0 for upright fonts.
* `font_is_embedded` — logical; `TRUE` if the font program is embedded.
* `font_flags` — PDF font-descriptor `/Flags` bitmask (fixed-pitch,
  serif, symbolic, ...).

## Per-object access

When you already have a specific text-typed `pdfium_obj` in hand —
typically from filtering `pdf_page_objects()` — three accessors give
you the same data piecewise:

```{r}
texts <- Filter(function(o) o$type == "text", pdf_page_objects(page))
length(texts)

pdf_text_content(texts[[1L]]) # the Unicode string
pdf_text_font_size(texts[[1L]]) # font size in user-space points
str(pdf_text_font(texts[[1L]])) # named list of font metadata
```

`pdf_text_runs()` is essentially a batched form of these three accessors
joined with `pdf_obj_bounds()`, returned as a single tibble.

## Unicode and font notes

UTF-16LE comes through correctly for every script PDFium knows how to
decode — including CJK ideographs, Latin diacritics, and characters
outside the Basic Multilingual Plane that need surrogate pairs:

```{r}
Encoding(pdf_text_content(texts[[1L]]))
nchar(pdf_text_content(texts[[1L]]))
```

A few caveats worth knowing about:

* **Glyph IDs vs Unicode.** If a PDF uses a non-standard font without a
  `/ToUnicode` CMap, PDFium can't always map glyph IDs back to Unicode.
  In those cases `pdf_text_content()` returns the best-effort decoding;
  some characters may come through as private-use codepoints or as `?`.
  This is a property of the source PDF, not of `pdfium`.
* **Text matrix.** `font_size_pt` is the font size that PDFium reports.
  The effective glyph height on the page is
  `font_size_pt * abs(matrix$d)` where `matrix` comes from
  `pdf_obj_matrix(text_obj)`.

## Combining text and path data

The `pdf_extract_paths()` one-call helper attaches `pdf_text_runs()`
output as a `text_runs` attribute on its result, so downstream
consumers that want to label paths with nearby text have everything
they need in one object. See the
[extracting-paths](extracting-paths.html) vignette for the schema.

## Cleanup

```{r}
pdf_page_close(page)
pdf_doc_close(doc)
```
